Historical Soil Organic Carbon Budget

### 2020-05-03 SOCbudgetPaper/SOCbudget.tex
methods update, add more result rmds

\begin{abstract}
SOC one of larges c sinks on earth (3 times larger biospehre pool).
Agricultural management leads to a depletion of soil organic crabon.
However this depletion of soil organic carbon (SOC) pools are so far not
well represented in global assessments of historic carbon emissions.
While SOC models often represent well the biochemical processes that
lead to the accumulation and decay of SOC, the management decisions
driving these biophysical processes are still little investigated. Here
we create a spatial explicit data set for crop residue and manure
management on cropland based on global historic production (FAOSTAT) and
land-use (LUH2) data and combine it with the IPCC Tier 2 approach to
create a half-degree resolution soil organic carbon budget on mineral
soils. We estimate that due to arable farming soils have lost over (?)
GtOC of which (??) GtOC have been released within the period 1990-2010.
Tier 2 IPCC methodolgy estimates higher soil organic carbon losses than
Tier 1 methods, which may origin from \ldots{} . We also find that SOC
is very sensity to management decision such as residue recycling
indicating the nessessity to incorporated better management data in soil
models.
\end{abstract}

@@

Following the tier 2 approach of the refinement of IPCC guidelines
vol.~4 (\citet{ipcc_2019_2019}), we calculate annual land use type
specific soil organic carbon stocks for cropland, pastures and natural
vegetation on half-degree resolution for the period of 1965 to 2010
based on the following three steps: (1) Calculating the land use
(sub-)type specific steady-states and decay rates for SOC stocks given
the current biophysical, climatic and agronomic conditions, (2)
accounting for land conversation effects by transferring SOC between
land use types and (3) updating SOC stocks based on the previous stock,
the steady-state and the decay rate.

@@

In a simple first order kinetic approach the steady-state soil organic
carbon stocks \(SOC^{eq}\) are given by \begin{equation}
SOC^{eq} =\frac{C^{\textrm{in}}}{k}
\label{eq:inoutflow}
\end{equation} with \(C^{\textrm{in}}\) being carbon inputs to the soil
and \(k\) denoting the soil organic carbon decay rate. We use for our
calculations the steady-state method of the refinement of the IPCC
guidelines vol.~4 (\citet{ipcc_2019_2019}) for mineral soils, which
assume three soil carbon sub-pools (active, slow and passive) and
entangled dynamics between them. Carbon inflow to each sub-pool (see
@ref(sec:carboninputs)) and decay rates (see @ref(sec:tier2)) of each
sub-pool are still the key components to determining steady-state SOC
stocks.

\hypertarget{sec:carboninputs}{%
\subsubsection{Carbon Inputs to the Soil}\label{sec:carboninputs}}

We account for different carbon input sources depending on the land use
type (see table @ref(tab:datasourceinputs)). Following the IPCC
methodology carbon inputs are disaggregated into metabolic and
structural components depending on their lignin and nitrogen content
(see @ref(ipcc\_2019\_2019)). For each component the sum over all carbon
input sources is allocated to the respective SOC sub-pools via transfer
coefficients. This implies that not only the amount of carbon, but also
their structural composition is determining the effective inflow. Data
sources for all considered carbon inputs as well as for lignin and
nitrogen content can be found in table @ref(tab:datasourceinputs).

 \begin{table*}[h]
 \caption{Type and data sources for carbon inputs to different land use types}
 \begin{tabular}{l l l l}
 \tophline
  Land use types   & source of carbon inputs & data source & nitrogen and lignin content \\
 \middlehline
 \multirow{3}{*}{Cropland} & residues & FAOSTAT, LPJmL4 [2, sec:residues] & default values given by [2]  \\
                            & dead below ground biomass of crops & FAOSTAT, LPJmL4 [2, sec:residues] & default values given by [2] \\
                            & manure & FAOSTAT, Isabelle [2, sec:manure] & default values given by [2] \\
                            \hline
 \multirow{2}{*}{Pasture}  & annual litterfall & LPJmL4 [3] & default values given by [2] \\ 
                            & manure  & FAOSTAT, Isabelle [2, sec:manure] & default values given by [2] \\
                            \hline
  Natural vegetation        & annual litterfall & LPJmL4 [4]& \begin{minipage}[t]{0.28\columnwidth}\raggedright\strut Nitrogen and lignin content of tree compartments used in CENTURY [4] \strut \end{minipage}\tabularnewline
 \bottomhline
 \end{tabular}
 \label{tab:datasourceinputs}
 \belowtable{}
 \end{table*}

\hypertarget{sec:tier2}{%
\subsubsection{Soil Organic Carbon decay (300)}\label{sec:tier2}}

The sub-pool specific decay rates are influenced by climatic conditions,
biophysical and biochemical soil properties as well as management
factors that vary over time (t) and space (i). Following the
steady-state method of the refinement of the IPCC guidelines vol.~4
(\citet{ipcc_2019_2019}) for mineral soils we consider temperature
(temp), water (wat), sand fraction (sf) and tillage (till) effects to
account for spatial variation of decay rates. Thus \(k_{sub}\) is given
by

@@

For cropland we distinguish the effect of different tillage (see
@ref(\#sec:tillage)) and irrigation (see @ref(\#sec:irrigation))
practices on decay rates, whereas on pastures and natural vegetation, we
assume rainfed and non-tilled conditions. Data sources as well as
considered effects for each land use types are shown in table
@ref(tab:datasourcedecay). To account for variations of decay rates
within each grid cell due to different tillage and irrigation regimes,
average rates based on area shares are calculated.

 \begin{table*}[h]
 \caption{Type and data sources for carbon inputs to different land use types}
 \begin{tabular}{l l l l}
 \tophline
  Land use types   & type of decay driver & parameter use to represent driver & data source \\
 \middlehline
 \multirow{2}{*}{all} & Soil quality & Sand fraction of the first 0-30 cm &  [SoilGrids]  \\
                      \cline{2-4}

@@

\subsubsection{SOC transfer between land use types}

We calculate SOC stocks based on the area shares of land use types (lut)
within the half-degree grid cells (i). If land is converted from one
land use type into others (!lut), the respective share of the SOC stocks
is reallocated. We account for land conversion at the beginning of each
time step \(t\) by calculating a preliminary stock \(SOC_{lut,t*}\) via

@@

with \(A\) being the area, \(AR\) the area reduction and \(AE\) the area
expansion for a given land use type \(lut\). Note that \(!lut\) denotes
the sum over all other land use types, which decreases in the specific
time step \(t\). Data sources and methodology on land use states and
changes are described in @ref(sec:landuse).

@@

\begin{equation}
SOC_{t} = \sum_{i} \underbrace{\sum_{lut} \overbrace{\sum_{sub} SOC_{lut, sub, t, i}.}^{\text{$SOC_{lut, t, i}$ - land use type specific SOC stock within cell}}}_{\text{$SOC_{t, i}$ - total SOC stock within cell}}
\label{eq:totalstock}
\end{equation}

@@

To initialize all SOC sub-pools we assume that

@@

@@

Land use patterns are based on the Land-Use Harmonization 2 (LUH2,
\citep{LUH2}) data set, which we aggregate from quarter degree to half
degree resolution. We disaggregate the five different cropland
subcategories (c3ann, c3per, c4ann, c4per, c3nfx) of LUH2 into our 17
crop groups, assuming relative shares for each gridcell based on the
country and year specific area shares of FAOSTAT data (\citep{FAOSTAT})
(see @ref(append:Table\_luh2fao2mag) for more details on the crop type
mapping). Land use transitions are calculated as net area differences of
the land use data on half-degree.

@@

@@

@@

A fixed share of the ag residues is assumed to be burned on field, which
depends on the per-capita income of the country. Following
\citep{smil1999}) we assume 25\% burn share for low-income countries
according to worldbank definitions (\(<\,1000\,\tfrac{USD}{yr}\)), 15\%
for high-income (\(>\,10000\,\tfrac{USD}{yr}\) and linearly interpolate
shares for all middle-income countries depending on their per-capita
income.

Residue demand for feed is based on country and livestock specific feed
basekts (see \citep{weindl}) taking available ag residual biomass as
well as livestock productivity into account. The three residue groups
(straw, high-lignin and low-lignin residues) of the feed baskets are
disaggregated to the crop groups, using\ldots. (see
@ref(append:Table\_kcr2kres)). \textless-- its a disggregation, no
--\textgreater{}

We estimated a material use share for the straw residues group of 5\%
and a fuel demand share demand of 10\% \textless-- only for straw, or
for all groups? --\textgreater{} in low income countries.

The remaining ag residues as well as all bg residues are assumend to be
recycled to the soil.

@@

@@

To disaggregate country level FAOSTAT livestock production values to
half-degree pattern, we use the following rule based assumptions which
were inspired by the approach of \citep{gilbert}.

For poultry, egg and monogastric meat production we divide into
intensive and extensive production systems based on a per-capita income
of the country. For low-income countries according to worldbank
definitions (\textless1000 USD/yr), we assume extensive production
systems. We located them according to (built-up areas shares\textbar{}
population shares) based on the idea that these animals are held in
households, subsistence or small-holder farming systems.

\begin{itemize}
\item
  Intensive production is distributed within a country using the crop
  production (excluding 2nd generation bioenergy crops) share, assuming
  that feed availability is the most driving factor for livestock
  location.
\item
  For dairy and ruminated meat production we
\item
  splot into pasture fed and non-pasture fed fractions based on the
  aggregated feed mix within a country (explain rules)
\end{itemize}

\begin{itemize}
\item
  Pasture fed animals are distributed to half-degree cells according to
  the pasture production shares of that cell within a country (not
  methodologically clear description).
\item
  Non-pasture fed animals are distributed again using crop production
  (excluding 2nd generation bioenergy crops) shares (not clear, either).
\end{itemize}

We calculate excretions by first estimating the nitrogen balance of the
livestock system (\citep{bodirsky1012}) on the basis of comprehensive
livestock feed baskets ({[}weindl{]}), assuming that all nitrogen in
protein feed intake, minus the nitrogen in the slaugher mass, is
excreted. Carbon in excreted manure is estimated by applying fixed C:N
ratios (given by \citep[(][]{ipcc_2019_2019}). What about animal waste
management losses?

>>>>

\begin{abstract}
SOC one of larges c sinks on earth (3 times larger biosphere pool).
Agricultural management leads to a depletion of soil organic carbon.
However this depletion of soil organic carbon (SOC) pools are so far not
well represented in global assessments of historic carbon emissions.
While SOC models often represent well the biochemical processes that
lead to the accumulation and decay of SOC, the management decisions
driving these biophysical processes are still little investigated. Here
we create a spatial explicit data set for crop residue and manure
management on cropland based on global historic production (FAOSTAT) and
land-use (LUH2) data and combine it with the IPCC Tier 2 approach to
create a half degree resolution soil organic carbon budget on mineral
soils. We estimate that due to arable farming soils have lost over (?)
GtOC of which (??) GtOC have been released within the period 1990-2010.
We show that, our results on global scale based on Tier 2 IPCC
methodolgy are in good agreement with Tier 1 default assumptions. We
also find that SOC is very sensitive to management decision such as
residue recycling indicating the nessessity to incorporated better
management data in soil models.
\end{abstract}

@@

Following the tier 2 approach of the refinement of IPCC guidelines
vol.~4 (\citet{ipcc_2019_2019}), we estimate global land-use type
specific soil organic carbon (SOC) stocks for cropland and natural
vegetation on half-degree resolution from 1965 to 2010. We assume the
actual SOC state converges towards a stable steady state, that itself is
changing over time and space depending on biophysical, climatic and
agronomic conditions. Therefor we conduct the following three steps: (1)
Calculating annual land-use (sub-)type specific steady states and decay
rates for SOC stocks, (2) accounting for land conversion by transferring
SOC between land-use types and (3) updating SOC stocks based on the
previous stock, the steady state and the decay rate.

@@

In a simple first order kinetic approach the steady-state soil organic
carbon stocks \(SOC^{eq}\) are given by \begin{equation}
SOC^{eq} =\frac{C^{\textrm{in}}}{k}
\label{eq:inoutflow}
\end{equation} with \(C^{\textrm{in}}\) being carbon inputs to the soil
and \(k\) denoting the soil organic carbon decay rate. We use for our
calculations the steady-state method of the refinement of the IPCC
guidelines vol.~4 (\citet{ipcc_2019_2019}) for mineral soils, which
assume three soil carbon sub-pools (active, slow and passive) and
entangled dynamics between them. Annual carbon inflow to each sub-pool
(see @ref(sec:carboninputs)) and annual decay rates (see
@ref(sec:tier2)) of each sub-pool are still the key components to
determining steady-state SOC stocks.

\hypertarget{sec:carboninputs}{%
\paragraph{Carbon Inputs to the Soil}\label{sec:carboninputs}}

We account for different carbon input sources depending on the land-use
type (see table @ref(tab:datasourceinputs)). Following the IPCC
methodology carbon inputs are disaggregated into metabolic and
structural components depending on their lignin and nitrogen content
(see @ref(ipcc\_2019\_2019)). For each component the sum over all carbon
input sources is allocated to the respective SOC sub-pools via transfer
coefficients. This implies that not only the amount of carbon, but also
their structural composition is determining the effective inflow. Data
sources for all considered carbon inputs as well as for lignin and
nitrogen content can be found in table @ref(tab:datasourceinputs).

 \begin{table*}[h]
 \caption{Type and data sources for carbon inputs to different land-use types}
 \begin{tabular}{l l l l}
 \tophline
  land-use types   & source of carbon inputs & data source & nitrogen and lignin content \\
 \middlehline
 \multirow{3}{*}{Cropland} & residues & FAOSTAT, LPJmL4 [2, sec:residues] & default values given by [2]  \\
                            & dead below ground biomass of crops & FAOSTAT, LPJmL4 [2, sec:residues] & default values given by [2] \\
                            & manure & FAOSTAT, Isabelle [2, sec:manure] & default values given by [2] \\
                            \hline
  Natural vegetation        & annual litterfall & LPJmL4 [4]& \begin{minipage}[t]{0.28\columnwidth}\raggedright\strut Nitrogen and lignin content of tree compartments used in CENTURY [4] \strut \end{minipage}\tabularnewline
 \bottomhline
 \end{tabular}
 \label{tab:datasourceinputs}
 \belowtable{}
 \end{table*}

\hypertarget{sec:tier2}{%
\paragraph{Soil Organic Carbon decay (300)}\label{sec:tier2}}

The sub-pool specific decay rates are influenced by climatic conditions,
biophysical and biochemical soil properties as well as management
factors that all vary over time (t) and space (i). Following the
steady-state method of the refinement of the IPCC guidelines vol.~4
(\citet{ipcc_2019_2019}) for mineral soils we consider temperature
(temp), water (wat), sand fraction (sf) and tillage (till) effects to
account for spatial variation of decay rates. Thus \(k_{sub}\) is given
by

@@

For cropland we distinguish the effect of different tillage (see
@ref(\#sec:tillage)) and irrigation (see @ref(\#sec:irrigation))
practices on decay rates, whereas on natural vegetation, we assume
rainfed and non-tilled conditions. Data sources as well as considered
effects for each land-use types are shown in table
@ref(tab:datasourcedecay). To account for variations of decay rates
within each grid cell due to different tillage and irrigation regimes,
average rates based on area shares are calculated.

 \begin{table*}[h]
 \caption{Type and data sources for carbon inputs to different land-use types}
 \begin{tabular}{l l l l}
 \tophline
  land-use types   & type of decay driver & parameter use to represent driver & data source \\
 \middlehline
 \multirow{2}{*}{all} & Soil quality & Sand fraction of the first 0-30 cm &  [SoilGrids]  \\
                      \cline{2-4}

@@

\subsubsection{SOC transfer between land-use types}

We calculate SOC stocks based on the area shares of land-use types (lut)
within the half-degree grid cells (i). If land is converted from one
land-use type into others (!lut), the respective share of the SOC stocks
is reallocated. We account for land conversion at the beginning of each
time step \(t\) by calculating a preliminary stock \(SOC_{lut,t*}\) via

@@

with \(A\) being the area, \(AR\) the area reduction and \(AE\) the area
expansion for a given land-use type \(lut\). Note that \(!lut\) denotes
the sum over all other land-use types, which decreases in the specific
time step \(t\). Data sources and methodology on land-use states and
changes are described in @ref(sec:landuse).

@@

\begin{equation}
SOC_{t} = \sum_{i} \underbrace{\sum_{lut} \overbrace{\sum_{sub} SOC_{lut, sub, t, i}.}^{\text{$SOC_{lut, t, i}$ - land-use type specific SOC stock within cell}}}_{\text{$SOC_{t, i}$ - total SOC stock within cell}}
\label{eq:totalstock}
\end{equation}

@@

To initialize all SOC sub-pools we assume that cropped land natural
vegetation * steady-states or * spin up

@@

Agricultural management data is based on R library package moinput
(\citep{bodirsky_in_prep}), which compiles in its core country-specific
FAO production and cropland statistics (\citep{FAOSTAT}) to a
compherensive and constistent data suite. The data is prepared in 5 year
time steps from 1965 to 2010, which also restricts our analysis to this
time span. For all the following data, if not declared differently, we
interpolate values linearly between the time steps and hold it constant
before the first time step for a short spin-up phase from 1961 to 1965.

@@

Land-use patterns are based on the Land-Use Harmonization 2 (LUH2,
\citep{LUH2}) data set, which we aggregate from quarter degree to half
degree resolution. We disaggregate the five different cropland
subcategories (c3ann, c3per, c4ann, c4per, c3nfx) of LUH2 into our 17
crop groups, assuming relative shares for each gridcell based on the
country and year specific area shares of FAOSTAT data (\citep{FAOSTAT})
(see @ref(append:Table\_luh2fao2mag) for more details on the crop type
mapping). land-use transitions are calculated as net area differences of
the land-use data on half-degree.

@@

\paragraph{Crop Production}

@@

\paragraph{Crop Residue Production}

@@

\subparagraph{Burned Residues}

A fixed share of the ag residues is assumed to be burned on field
depending on the per-capita income of the country. Following
\citep{smil1999}) we assume 25\% burn share for low-income countries
according to worldbank definitions (\(<\,1000\,\tfrac{USD}{yr}\)), 15\%
for high-income (\(>\,10000\,\tfrac{USD}{yr}\) and linearly interpolate
shares for all middle-income countries depending on their per-capita
income. Depending on the crop type 80--90\% of the residue carbon burned
on the fields are lost within the combustion process
(\citep{ipcc_2006_2006}).

\subparagraph{Residue Usage}

We compile out of our 17 crop groups, three used residue groups (straw,
high-lignin and low-lignin residues) with additional demand for other
purposes and one residues with no double use (see
@ref(append:Table\_kcr2kres)).

\subparagraph{Feed Demand}

Residue feed demand for five different livestock groups is based on
country- and residue-group-specific feed basekts (see \citep{weindl})
taking available ag residual biomass as well as livestock productivity
into account.

\subparagraph{Material and Fuel demand}

We estimate a material-use share for the straw residues group of 5\% and
a fuel-share of 10\% for all used residues groups in low income
countries according to worldbank definitions
(\(<\,1000\,\tfrac{USD}{yr}\)). For high-income
(\(>\,10000\,\tfrac{USD}{yr}\) no withdrawl for material or fuel use is
assumend, leaving middle-income countries with linearly interpolate
shares depending on their per-capita income.

\subparagraph{Recycled Residues}

The remaining ag residues as well as all bg residues are assumend to be
recycled to the soil. We cut high recycling shares per hectar at the
95\%-percentile to corrected for outliers.

\paragraph{Pasture Production}

@@

\paragraph{Dry Matter to Carbon Transformation}

@@

\paragraph{Livestock Distribution}

To disaggregate country level FAOSTAT livestock production values to
half-degree pattern, we use the following rule based assumptions which
were inspired by the approach of \citep{gilbert} and uses feed basket
assumptions based on a revised methodology of \citet{weindl}.

\subparagraph{Livestock Managament Intensities}

To account for different livestock management intensities, we
differentiate between extensive and an intensive systems assuming that
they have different requirements on production factors such as feed
availabiliy, access to markets and infrastructure (e.g.~roads, water).

\subparagraph{Dairy and ruminant Meat Production}

For extensive dairy and ruminant meat production we estimate, that
livestock is located rather immovable on pastures and rangeland. We use
the country-level pasture feed stuff shares within the feed baskets, to
split up pasture-fed, extensive from the rather intensive, crop-fed
livestock. Extensive production of dairy and ruminant meat is allocated
weighted by the pasture production values to all half-degree gridcell of
a country. On the other hand intensive dairy and ruminant meat
production is assumend to be located proportinal to crop production to
have short transport distances for feed stuff.

\subparagraph{Poultry, Egg and monogastric Meat Production}

For poultry, egg and monogastric meat production we use the per-capita
income of the country to divide into intensive and extensive production
systems. For low-income countries according to worldbank definitions
(\textless1000 USD/yr), we assume extensive production systems. We
located them according to built-up areas shares based on the idea that
these animals are held in households, subsistence or small-holder
farming systems with a high labour per animal ratio. Intensive
production is distributed within a country using the crop production
share, assuming that feed availability is the most driving factor for
livestock location.

\paragraph{Manure Excretion}

Manure production and management is based on a revised methodology of
(\citep{bodirsky2012}) and will be explained in key aspects again due to
its central role for soil carbon modelling. Based on the gridded
livestock distribution we calculate excretions by estimating the
nitrogen balance of the livestock system on the basis of comprehensive
livestock feed baskets (\citep{weindl}), assuming that all nitrogen in
protein feed intake, minus the nitrogen in the slaughter mass, is
excreted. Carbon in excreted manure is estimated by applying fixed C:N
ratios (given by \citep[(][]{ipcc_2019_2019}).

\paragraph{Manure Storage}

Depending on the feed system we assume manure to be handled in four
different ways: All manure orginated from pasture feed intake is
excreted directly to pastures and rangelands (pasture grazing) deducting
manure collected as fuel. Manure fuel shares are estimated using IPCC
default values (\citet{ippc_2006_2006}). Whereas for low-income
countries according to worldbank definitions (\textless1000 USD/yr), we
adopt a share of 25\% of crop residues in feed intake directly consumend
and excreted on crop fields (stubble grazing), we do not consider any
stubble grazing in high-income countries (\(>\,10000\,\tfrac{USD}{yr}\),
leaving middle-income countries with linearly interpolate shares
depending on their per-capita income. For all other feed items we assume
the manure to be stored in animal waste management systems associated to
animal houses.

\paragraph{Manure Recycling}

To estimate the carbon actually recycled to the soil, we account for
carbon losses during storage and recycling shares in different animal
waste management and grazing systems. Whereas we assume no losses for
pasture and stubble grazing, we consider all carbon lost for manure
collected as fuel. For manure stored in different animal waste
management system we compiled carbon loss rates partly depending on
nitrogen loss rates as specified in \citep{bodirsky2012} (see
@ref(append:Table\_clossAWMS))).
